---
title: "21966155 - A3"
output: html_notebook
---

*) Loading all needed package:
```{r}
library(readr)
library(readxl)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(stringr)
#install.packages("Hmisc")
library(Hmisc)
```

TASK 1:
```{r}
employee_demographics <- read.csv("survey_employee_demographics.csv", na.strings = c("", "-", "--"))
employee_employment_details <- read_excel("survey_employee_employment_details.xlsx", na = c("", "-", "--"))
combined_employee <- merge(employee_demographics, employee_employment_details, by = "RecordID")
combined_employee
```

TASK 2:
  Task 2.a:
```{r}
describe(combined_employee)
summary(combined_employee)
```
  
  Task 2.c:
  +) Change the name of data column and put them in the new dataframe. The new dataframe is "clean_combined_employee". This datafarme will also be used in the rest of the questions.
```{r}
clean_combined_employee <- combined_employee %>%
  rename(
    `Total Years of Experience` = TotalYearsOfExperience,
    `Years of Experience in Germany` = YearsOfExperienceInGermany,
    `Main Programming Language` = YourMainProgrammingLanguage,
    `Other Programming Language` = OtherProgrammingLanguagesYouUseOften,
    `Yearly Brutto Salary (EUR)` = YearlyBruttoSalaryInEUR,
    `Yearly Bonus and Stocks (EUR)` = YearlyBonusAndStocksInEUR,
    `Annual Brutto Salary Last Year` = AnnualBruttoSalaryOneYearAgo,
    `Annual Bonus And Stocks Last Year` = AnnualBonusAndStocksOneYearAgo,
    `Number of Vacation Days` = NumberOfVacationDays,
    `Main Language At Work` = MainLanguageAtWork,
    `Lost Job due to COVID19` = HaveYouLostYourJobDueToTheCoronavirusOutbreak,
    `Kurzarbeit's Hours per Week` = HaveYouBeenForcedToHaveAShorterWorkingWeekKnownAsKurzarbeitIfYesHowManyHoursPerWeek,
    `WFH Additional Monetary from Employer (2020, EUR)` = HaveYouReceivedAdditionalMonetarySupportFromYourEmployerDueToWorkFromHomeIfYesHowMuchIn2020InEUR
  )
clean_combined_employee
```

  +) Removing rows that have too many NAs (more than 5 - 25% of the total columns in the dataset), except the last 2 rows as they depend on the Yes/No answer in the column `Lost Job due to COVID19`:
```{r}
na_counts <- rowSums(is.na(clean_combined_employee[ , !(names(clean_combined_employee) %in% c("Kurzarbeit's Hours per Week", "WFH Additional Monetary from Employer (2020, EUR)"))]))
clean_combined_employee <- clean_combined_employee[na_counts <= 5, ]
```

  +) Fixing inconsistency in the column "Total Years of Experience":
```{r}
#Modifying data entries:
  ##Converting comma to decimal point in some data entries:
clean_combined_employee$`Total Years of Experience` <- gsub(",", ".", clean_combined_employee$`Total Years of Experience`)

#Extracting the first numeric part of some data entries:
clean_combined_employee$`Total Years of Experience` <- str_extract(clean_combined_employee$`Total Years of Experience`, "\\d+\\.?\\d*")

#Assigning a numeric value to data entry "less than year":
clean_combined_employee$`Total Years of Experience`[clean_combined_employee$`Total Years of Experience` == "less than year"] <- "0.5" #assuming half a year for "less than a year"

#Double check to ensure that all of entries are in the correct format:
unique(clean_combined_employee$`Total Years of Experience`)
```
 
 +) Fixing inconsistency in the column "Years of Experience in Germany":
```{r}
#Modifying:
 ##Replacing comma in data entries
clean_combined_employee$`Years of Experience in Germany` <- gsub(",", ".", clean_combined_employee$`Years of Experience in Germany`)
 ##Transferring text to numeric
clean_combined_employee <- clean_combined_employee %>%
  mutate (
    `Years of Experience in Germany` = case_when(
      `Years of Experience in Germany` == "⁰" ~ "0",
      grepl("in Poland|in Switzerland", `Years of Experience in Germany`, ignore.case = TRUE) ~ "0",
      grepl("less than year", `Years of Experience in Germany`, ignore.case = TRUE) ~ "0.5",
      grepl("4 month", `Years of Experience in Germany`, ignore.case = TRUE) ~ "0.3",
      grepl("3 month", `Years of Experience in Germany`, ignore.case = TRUE) ~ "0.25",
      grepl("<1", `Years of Experience in Germany`, ignore.case = TRUE) ~ "0.8",
      grepl("lab",`Years of Experience in Germany`, ignore.case = TRUE) ~ "0",
      TRUE ~ `Years of Experience in Germany`
    )
  )
#Double check to ensure that all of entries are in the correct format:
unique(clean_combined_employee$`Years of Experience in Germany`)
```
  
  +) Fixing inconsistency in the column "SeniorityLevel":
```{r}
#Standardizing data entries:
clean_combined_employee <- clean_combined_employee %>%
  mutate(
    SeniorityLevel = case_when(
      grepl("intern|student", SeniorityLevel, ignore.case = TRUE) ~ "Intern/Student",
      grepl("junior", SeniorityLevel, ignore.case = TRUE) ~ "Junior",
      grepl("senior", SeniorityLevel, ignore.case = TRUE) ~ "Senior",
      grepl("lead|key|principal", SeniorityLevel, ignore.case = TRUE) ~ "Lead",
      grepl("entry level", SeniorityLevel, ignore.case = TRUE) ~ "Junior",
      grepl("head|Work Center", SeniorityLevel, ignore.case = TRUE) ~ "Manager", 
      grepl("cto|c-level|c level|director|vp", SeniorityLevel, ignore.case = TRUE) ~ "C Level",
      grepl("no level|no idea", SeniorityLevel, ignore.case = TRUE) ~ NA,
      TRUE ~ SeniorityLevel
    ),
    SeniorityLevel = str_to_title(SeniorityLevel)  
  )

# Printing the updated seniority levels to verify the changes
unique(clean_combined_employee$SeniorityLevel)
```

  +) Fixing format column "Yearly Brutto Salary (EUR)":
```{r}
#Removing decimals from data entries
clean_combined_employee$`Yearly Brutto Salary (EUR)` <- round(as.numeric(clean_combined_employee$`Yearly Brutto Salary (EUR)`))

#Replacing outliers with NAs
salary_outliers <- boxplot.stats(clean_combined_employee$`Yearly Brutto Salary (EUR)`)$out
salary_outliers

which(clean_combined_employee$`Yearly Brutto Salary (EUR)` %in% salary_outliers)

clean_combined_employee[which(clean_combined_employee$`Yearly Brutto Salary (EUR)` %in% salary_outliers), "Yearly Brutto Salary (EUR)"] <- NA

##Double check to ensure that all of entries are in the correct format
summary(clean_combined_employee$`Yearly Brutto Salary (EUR)`)

```
  
  +) Fixing inconsistency in the column "EmploymentStatus":
```{r}
#Standardizing:
clean_combined_employee <- clean_combined_employee %>%
  mutate(EmploymentStatus = case_when(
    grepl("working|student|intern", EmploymentStatus, ignore.case = TRUE) ~ "Intern",
    EmploymentStatus == "Full-time position, part-time position, & self-employed (freelancing, tutoring)" ~ "Mixed",
    EmploymentStatus == "full-time, but 32 hours per week (it was my request, I'm a student)" ~ "Full-time employee",
    EmploymentStatus == "Full-time position, part-time position, & self-employed (freelancing, tutoring)" ~ "Mixed",
    TRUE ~ EmploymentStatus
  ))
#Double check to ensure that all of entries are in the correct format
unique(clean_combined_employee$EmploymentStatus)
```

  +) Fixing inconsistency in the column "Position":
```{r}
#Standardizing:
clean_combined_employee <- clean_combined_employee %>%
  mutate(
    Position = case_when(
       Position == "working student (QA)" ~ "QA Engineer",
       Position == "Testautomation" ~ "Test Automation",
       Position == "Testmanager" ~ "Test Manager",
       Position == "Systemadministrator" ~ "System Administrator", 
       Position == "Software Architekt" ~ "Software Architect",
       Position == "Senior Network&Security Eng." ~ "Security Engineer", 
       grepl("dana|dba", Position, ignore.case = TRUE) ~ "Data Analyst",
       grepl("sre|sw", Position, ignore.case = TRUE) ~ "Software Engineer",
       grepl("consult", Position, ignore.case = TRUE) ~ "Consultant",
       grepl("sales", Position, ignore.case = TRUE) ~ "Sales", 
       grepl("Fullstack Engineer, Ну Или Software Engineer", Position, ignore.case = TRUE) ~ "Fullstack Engineer", 
       grepl("analyst", Position, ignore.case = TRUE) & !grepl("data analyst", Position, ignore.case = TRUE) ~ "Business Analyst",
       grepl("Databengineer", Position, ignore.case = TRUE) ~ "Data Engineer",
       grepl("(?=.*\\bdata\\b)(?=.*\\bengineer\\b)", Position, ignore.case = TRUE, perl = TRUE) ~ "Data Engineer",
       grepl("(?=.*\\bsoftware\\b)(?=.*\\bengineer\\b)", Position, ignore.case = TRUE, perl = TRUE) ~ "Software Engineer",
       grepl("tech.*lead|technical lead", Position, ignore.case = TRUE) ~ "Tech Lead",
       grepl("teamlead|team lead", Position, ignore.case = TRUE) ~ "Team Lead",
       grepl("Full-Stack", Position, ignore.case = TRUE) ~ "Fullstack Developer",
       TRUE ~ (Position),
    ),
    Position = str_to_title(str_trim(Position))
  )
unique(clean_combined_employee$Position)   
```
  +) Fixing inconsistency in column "Lost Job due to COVID19":
```{r}
#Modifying inconsistency:
clean_combined_employee <- clean_combined_employee %>%
  mutate(
     `Lost Job due to COVID19`= case_when(
       `Lost Job due to COVID19` == "i didn't but will be looking for new one because of covid" ~ "No",
       `Lost Job due to COVID19` == "Lost the job but for different reason" ~ "No",
       `Lost Job due to COVID19` == "Have been a freelancer at the beginning of year" ~ "No",
       `Lost Job due to COVID19` == "Leads and project inquiries have slowed down" ~ "No",
       `Lost Job due to COVID19` == "Laid off for a bit" ~ "No",
       grepl("No", `Lost Job due to COVID19`, ignore.case = TRUE) ~ "No",
       grepl("kurzarbeitzeit", `Lost Job due to COVID19`, ignore.case = TRUE) ~ "No",
       grepl("Yes", `Lost Job due to COVID19`, ignore.case = TRUE) ~ "Yes",
       TRUE ~ (`Lost Job due to COVID19`),
    ),
   `Lost Job due to COVID19` = str_to_title(str_trim(`Lost Job due to COVID19`))
  )
unique(clean_combined_employee$`Lost Job due to COVID19`) 
```

  +) Change type of data in each column:
```{r}
clean_combined_employee <- clean_combined_employee %>%
  mutate(
    Gender = as.factor(Gender),
    Age = as.numeric(Age),
    `Total Years of Experience` = as.numeric(`Total Years of Experience`),
    `Years of Experience in Germany` = as.numeric(`Years of Experience in Germany`),
    `Number of Vacation Days` = as.numeric(`Number of Vacation Days`),
    `Yearly Brutto Salary (EUR)` = as.numeric(`Yearly Brutto Salary (EUR)`),
    `Yearly Bonus and Stocks (EUR)` = as.numeric(`Yearly Bonus and Stocks (EUR)`),
    `Annual Bonus And Stocks Last Year` = as.numeric(`Annual Bonus And Stocks Last Year`),
    `Annual Brutto Salary Last Year` = as.numeric(`Annual Brutto Salary Last Year`),
    `Kurzarbeit's Hours per Week` = as.numeric(`Kurzarbeit's Hours per Week`),
    `WFH Additional Monetary from Employer (2020, EUR)` = as.numeric(`WFH Additional Monetary from Employer (2020, EUR)`),
    EmploymentStatus = as.character(EmploymentStatus),
    `Lost Job due to COVID19` = as.factor(`Lost Job due to COVID19`)
  )

```

TASK 3:
  +) Histogram:
```{r}
breaks <- seq(26000, 110000, by = 10000)

ggplot(clean_combined_employee, aes(x = `Yearly Brutto Salary (EUR)`)) +
  geom_histogram(aes(y = ..count../sum(..count..)), breaks = breaks, fill = "lightblue", color = "black", na.rm = TRUE) +
  geom_text(aes(y = ..count../sum(..count..), label = sprintf("%.2f%%", ..count../sum(..count..) * 100)), 
            stat = "bin", vjust = -0.5, color = "black", size = 3, breaks = breaks) +
  scale_x_continuous(breaks = breaks) +
  labs(title = "Histogram of Yearly Gross Salaries of IT Professionals",
       x = "Yearly Gross Salary (EUR)",
       y = "Percentage of Total") +
  theme_minimal() 

```

  +) Box plot:
```{r}

#Calculating statistics
stats <- boxplot.stats(clean_combined_employee$`Yearly Brutto Salary (EUR)`)$stats
stats_df <- data.frame(
  Value = stats)

#Plotting
ggplot(data = clean_combined_employee, aes(x = factor(1), y = `Yearly Brutto Salary (EUR)`)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  geom_text(data = stats_df, aes(x = 1, y = Value, label = stats), size = 3, vjust = -0.5, hjust = -0.1) +
  labs(title = "Boxplot of Yearly Gross Salaries of IT Professionals",
       x = "",  
       y = "Yearly Gross Salary (EUR)") +
  scale_x_discrete(labels = NULL) +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)

```

TASK 4:
```{r}
#Creating a new dataframe to prepare data for plotting
task_4_data <- clean_combined_employee %>%
  filter(!is.na(Gender), Gender != "Diverse") %>%  #Removing NAs and Diverse as Diverse has limited records
  group_by(Gender) %>%
  summarise(Stats = list(boxplot.stats(`Yearly Brutto Salary (EUR)`)$stats)) %>%
  unnest_wider(Stats, names_sep = "_") %>%
  pivot_longer(cols = starts_with("Stats_"), names_to = "Statistic", values_to = "Value", names_prefix = "Stats_") %>%
  mutate(Statistic = factor(Statistic, levels = c("min", "q1", "median", "q3", "max")))

# Plotting
ggplot(clean_combined_employee, aes(x = Gender, y = `Yearly Brutto Salary (EUR)`)) +
  geom_boxplot(aes(fill = Gender), color = "black") +
  geom_text(data = task_4_data, aes(x = Gender, y = Value, label = round(Value, 2)),
            position = position_dodge(width = 0.75), vjust = -0.5, hjust = 1, size = 3, color = "black") +
  labs(title = "Box Plot of Yearly Gross Salary by Gender",
       x = "Gender",
       y = "Yearly Gross Salary (EUR)") +
  scale_x_discrete(limits = unique(task_4_data$Gender)) +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)

```


TASK 5:
  1.Relationship between salary and total years of experience
```{r}
#Preparing data to plot the scatter plot:
task5_data_exp_year <- clean_combined_employee %>%
  filter(!is.na(`Total Years of Experience`), !is.na(`Yearly Brutto Salary (EUR)`))

#Building a linear model for the scatter plot:
model <- lm(`Yearly Brutto Salary (EUR)` ~ `Total Years of Experience`, data = task5_data_exp_year)

#Getting r square
model_summary <- broom::glance(model)
r_squared <- model_summary$r.squared  

#Plotting
ggplot(task5_data_exp_year, aes(x = `Total Years of Experience`, y = `Yearly Brutto Salary (EUR)`)) +
  geom_point() + 
  geom_smooth(method = "lm", color = "red") +
  annotate("text", x = Inf, y = Inf, label = sprintf("R^2 = %.2f", r_squared), hjust = 3, vjust = 7, color = "black", size = 4) +
  labs(title = "Scatter Plot of Total Years of Experience vs Yearly Gross Salary",
       x = "Total Years of Experience",
       y = "Yearly Gross Salary (EUR)") +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)

```

  2. Relationship between salary and employment status
```{r}
#Preparing data to plot
task5_data_emp_status <- clean_combined_employee %>%
  filter(!is.na(EmploymentStatus), !is.na(`Yearly Brutto Salary (EUR)`))

#Plotting:
ggplot(task5_data_emp_status, aes(x = EmploymentStatus, y = `Yearly Brutto Salary (EUR)`, fill = EmploymentStatus)) +
  geom_boxplot() +
  labs(title = "Box Plot of Yearly Gross Salary by Employment Status",
       x = "Employment Status",
       y = "Yearly Gross Salary (EUR)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.75)) +
  scale_y_continuous(labels = scales::comma)

```

  3. Relationship between salary and company size
```{r}
task5_data_comp_size <- clean_combined_employee %>%
  filter(!is.na(CompanySize))

ggplot(task5_data_comp_size, aes(x = CompanySize, y = `Yearly Brutto Salary (EUR)`)) +
  geom_boxplot(aes(fill = CompanySize), color = "black", na.rm = TRUE) +
  labs(title = "Box Plot of Yearly Gross Salary by Company Size",
       x = "Company Size",
       y = "Yearly Gross Salary (EUR)") +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)
```

  4. Relationship between salary and position
```{r}
salary_by_position <- clean_combined_employee %>%
  group_by(Position) %>%
  filter(!is.na(Position), !is.na(`Yearly Brutto Salary (EUR)`)) %>%
  summarise(Count = n(), AverageSalary = round(mean(`Yearly Brutto Salary (EUR)`, na.rm = TRUE))) %>%
  arrange(desc(AverageSalary)) %>%
  filter(Count > 2) %>% #removing the variables have less than 2 counts to remove noise in analysis, as they have too limited records for effective analysis
  slice_head(n = 10)

# Create a bar chart
ggplot(salary_by_position, aes(x = reorder(Position, AverageSalary), y = AverageSalary)) +
  geom_bar(stat = "identity", fill = "pink") +
  labs(title = "Top 10 Average Yearly Gross Salary by Position",
       x = "Position",
       y = "Average Yearly Gross Salary (EUR)") +
  theme_minimal() +
  geom_text(aes(label = scales::comma(AverageSalary)), vjust = -0.5, color = "black", size = 2.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

  5. Relationship between salary and seniority level: 
```{r}
salary_by_seniority <- clean_combined_employee %>%
  group_by(SeniorityLevel) %>%
  filter(!is.na(SeniorityLevel), !is.na(`Yearly Brutto Salary (EUR)`)) %>%
  dplyr::summarize(AverageSalary = mean(`Yearly Brutto Salary (EUR)`, na.rm = TRUE)) %>%
  arrange(desc(AverageSalary))

#Plotting:
ggplot(salary_by_seniority, aes(x = reorder(SeniorityLevel, AverageSalary), y = AverageSalary)) +
  geom_bar(stat = "identity", fill = "pink") +
  labs(title = "Average Salaries by Seniority Level",
       x = "Seniority Level",
       y = "Average Salary (EUR)") +
  theme_minimal() +
  geom_text(aes(label = scales::comma(AverageSalary)), vjust = -0.3, color = "black", size = 2.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```





 








